Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die Einhaltung der Echtzeit

### 2024-06-04 Masterarbeit_Robotik_Pamuk.tex
[DRAFT] chapter 4 finished

% Die nachfolgenden Pakete stellen sonst nicht benötigte Features zur Verfügung
\usepackage{blindtext}
%
% Einträge für Deckblatt, Kurzfassung, etc.
%
\title{Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die
Einhaltung der Echtzeit}
\author{Halil Pamuk, BSc}
\studentnumber{51842568}
%\author{Titel Vorname Name, Titel\and{}Titel Vorname Name, Titel}
%\studentnumber{XXXXXXXXXXXXXXX\and{}XXXXXXXXXXXXXXX}
\supervisor{Sebastian Rauh, MSc. BEng}
%\supervisor[Begutachter]{Titel Vorname Name, Titel}
%\supervisor[Begutachterin]{Titel Vorname Name, Titel}
%\secondsupervisor{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachter]{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachterinnen]{Titel Vorname Name, Titel}
\place{Wien}
\kurzfassung{Erstellung einer Echtzeit-Robotersteuerungsplattform unter Verwendung von Salamander OS, Xenomai, QEMU
und PCV-521 in der Yocto-Umgebung. Die Plattform basiert auf Salamander OS und nutzt Xenomai für Echtzeit-
Funktionen. Dazu muss im ersten Schritt die Virtualisierungsplattform evaluiert werden. (QEMU, Hyper-V, Virtual
Box, etc.) Als weiterer Schritt folgt die Anbindung eines Roboters über eine VARAN-Bus Schnittstelle. Das
gesamte System wird in der Yocto-Umgebung erstellt und konfiguriert.
Das Hauptziel der Arbeit ist es, herauszufinden, wie die Integration von Echtzeit-Funktionen und effizienten
Kommunikationssystemen in eine Robotersteuerungsplattform die Reaktionszeit und Zuverlässigkeit von
Roboteranwendungen verbessern kann}
\schlagworte{Schlagwort1, Schlagwort2, Schlagwort3, Schlagwort4}
\outline{Sections \ref{sec:salamander4-bare-metal} and \ref{sec:salamander4-virtualisation}  demonstrate the inital real-time latency values gathered for bare metal and virtualisation. 
}
\keywords{Echtzeit, Virtualisierung, Xenomai, VARAN}
%\acknowledgements{\blindtext}

@@

First, a comprehensive literature review is conducted to understand the current trends and challenges in real-time robot control. Based on the literature study, a suitable virtualisation platform is selected.

After the selection of the virtualisation platform, the robot control platform is implemented. This step includes the installation and configuration of Salamander OS, Xenomai, QEMU and PCV-521 in the Yocto environment. Once the platform has been implemented, the robot is connected via a VARAN bus interface. Finally, the platform is evaluated to determine how the integration of real-time functions and efficient communication systems improves the response time and reliability of robot applications.

@@

\clearpage
\section{Xenomai}
\noindent Xenomai 3 \cite{XenomaiXenomai} is a real-time framework that offers two paths to real-time performance. The first approach supplements the Linux kernel with a compact real-time core dubbed Cobalt, demonstrated in Figure \ref{fig:cobalt}. Cobalt runs side-by-side with Linux, but it handles all time-critical activities like interrupt processing and real-time thread scheduling with higher priority than the regular kernel activities.

@@

As a starting point, initial latency values of both the bare metal and virtualization versions were measured with the latency tool of the xenomai tool suite. Salamander 4 Bare Metal refers to the proprietary hardware of SIGMATEK used to employ the custom operating system. Salamander 4 virtualisation refers to a virtual version of the Salamander 4 hardware platform, achieved through QEMU/KVM. Sections \ref{sec:salamander4-bare-metal} and \ref{sec:salamander4-virtualisation} specify the details of the measurements for both versions. In the further course, the aim was to bring the latency values of the virtualization closer to those of the hardware.

@@

\noindent As a reference point, the latency program was executed on Salamander 4 bare metal for a duration of 10 minutes. The complete command used was \texttt{latency -h -s -T 600}. Figure \ref{fig:max_latency_hardware} shows the gathered latency values in microseconds.

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/max_latency_hardware.png}
	\caption[Latency in hardware]{Latency in hardware}
	\label{fig:max_latency_hardware}
\end{figure}

Figure \ref{fig:gnuplot_max_latency_hardware} depicts the variation in latency over the course of said time.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/0hardware/gnuplot_max_latency_hardware.png}
	\caption[Variation in latency of hardware]{Variation in latency of hardware}
	\label{fig:gnuplot_max_latency_hardware}
\end{figure}

@@

\begin{table}[h]
	\centering
	\caption{Latency Statistics in microseconds}
	\label{tab:latency_statistics}
	\begin{tabular}{|c|c|}\hline
	\textbf{Statistic} & \textbf{Value (in \textmu s)} \\\hline
	Average Latency & 4.06 \\\hline
	Maximum Latency & 10.71 \\\hline
	Minimum Latency & 2.82 \\\hline
	Standard Deviation & 0.85 \\\hline
	\end{tabular}
	\end{table}

\section{Salamander 4 Virtualisation}\label{sec:salamander4-virtualisation}
In addition to providing Salamander 4 on its own hardware, SIGMATEK has also developed a virtualised version of this operating system. It was developped using Yocto, an open source project that allows customised Linux distributions to be created for embedded systems \cite{WelcomeYoctoProject}. The virtualisation runs in a QEMU environment, which is an open source tool for hardware virtualisation \cite{QEMU}. With the help of the script depicted in code \ref{script:qemu_def}, Salamander 4 is started together with the necessary hardware components in the QEMU environment. This makes it possible to run Salamander 4 on a variety of host systems, regardless of the specific hardware of the host. Upon generating the necessary files, Yocto generates a QEMU folder with the following components shown in code \ref{lst:ls}.

@@

Here is a description of the used components:
\begin{itemize}
	\item \textbf{bzImage}: Compressed Linux kernel image, loaded by QEMU at system start.
	\item \textbf{ovmf.code.qcow2}: Firmware file for QEMU, enables UEFI boot process.
	\item \textbf{qemu\_def.sh}: Shell script, starts QEMU with correct parameters to boot Salamamder 4 OS.
	\item \textbf{stek-drive-c-image-sigmatek-core2.tar.gz}: Archive containing files for C drive, unpacked and copied to drive-c/ directory by qemu\_def.sh script.
	\item \textbf{drive-c}: Directory serving as C drive for QEMU system, created and filled by qemu\_def.sh script.
	\item \textbf{salamander-image-sigmatek-core2.ext4}: Root file system for Salamander 4 OS, used as hard drive for QEMU system.
	\item \textbf{vmlinux}: Uncompressed Linux kernel image, typically used for debugging, contains debugging symbols not present in bzImage.
\end{itemize}

\noindent The inital QEMU script after the custom Yocto build and the starting point for this work is shown in Code \ref{script:qemu_def}. This script is used to start QEMU with correct parameters to boot Salamamder 4 OS. It will be adjusted in chapter \ref{cha:real-time_tuning} in order to accompany real-time performance tunings.

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualisation},label={script:qemu_def}]
    #!/bin/sh

@@

    exec  qemu-system-x86_64 -M pc,accel=kvm -kernel ./bzImage \
    -m 2048 -drive file=salamander-image-sigmatek-core2.ext4,format=raw,media=disk \
    -append "console=ttyS0 console=tty1 root=/dev/sda rw panic=1 sigmatek_lrt.QEMU=1 ip=dhcp rootfstype=ext4 schedstats=enable" \
    -net nic,model=e1000,netdev=e1000 -netdev bridge,id=e1000,br=nm-bridge \
    -fsdev local,security_model=none,id=fsdev0,path=drive-c -device virtio-9p-pci,id=fs0,fsdev=fsdev0,mount_tag=/mnt/drive-C \
    -device vhost-vsock-pci,guest-cid=3,id=vsock0 \
    -drive if=pflash,format=qcow2,file=ovmf.code.qcow2 \
    -no-reboot -nographic
\end{lstlisting}
\end{minipage}

\noindent The script is run on a generic Ubuntu 22.04.4 system. The kernel version and other details are presented in Code \ref{output:uname_a_ub}, using the \texttt{uname -a} command.

@@

Figure \ref{fig:gnuplot_max_latency_default} depicts the variation in latency with default settings over the course of said time.

@@

\begin{table}[h]
	\centering
	\caption{Latency Statistics default in microseconds}
	\label{tab:x}
	\begin{tabular}{|c|c|}\hline
	\textbf{Statistic} & \textbf{Value (in \textmu s)} \\\hline
	Average Latency & 46.22us \\\hline
	Maximum Latency & 707.62us \\\hline
	Minimum Latency & 15.59us \\\hline
	Standard Deviation & 52.13us \\\hline
	\end{tabular}
	\end{table}
\clearpage

\chapter{Real-Time Performance Tuning}\label{cha:real-time_tuning}
\section{BIOS Configurations}
BIOS is firmware used to perform hardware initialization during the booting process. Some BIOS settings can have a significant impact on the real-time performance of the system [2]. The configurations in Table 4 are tuned for real-time performance.

@@

\clearpage
\section{Guest OS Configurations}
\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualisation},label={script:qemu_configuration_optimizations}]
    #!/bin/sh

@@

\chapter{To Include}\label{cha:to_include}
\section{Latency Comparison}
In the initial phase, a comparative latency analysis was conducted between the hardware version and the virtualized version of Salamander 4. For this purpose, the latency tool of the Xenomai test suite was used. The latency was measured under two conditions, idle and CPU-stressed. The goal was to optimize the latency of the virtualisation of Salamander 4 OS to closely match that of the bare metal version.

@@

Figure \ref{fig:kvm_exit_vapic} shows \texttt{kvm\_exit} frequency with APIC virtualisation. Comparing this to the previous Figures \ref{fig:kvm_exit_taskset} and \ref{fig:kvm_exits_default}, it can be observed that an APIC\_Write no longer occurs.

>>>>

% Die nachfolgenden Pakete stellen sonst nicht benötigte Features zur Verfügung
\usepackage{blindtext}
%
% Einträge für Deckblatt, Kurzfassung, etc.
%
\title{Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die
Einhaltung der Echtzeit}
\author{Halil Pamuk, BSc}
\studentnumber{51842568}
%\author{Titel Vorname Name, Titel\and{}Titel Vorname Name, Titel}
%\studentnumber{XXXXXXXXXXXXXXX\and{}XXXXXXXXXXXXXXX}
\supervisor{Sebastian Rauh, MSc. BEng}
%\supervisor[Begutachter]{Titel Vorname Name, Titel}
%\supervisor[Begutachterin]{Titel Vorname Name, Titel}
%\secondsupervisor{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachter]{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachterinnen]{Titel Vorname Name, Titel}
\place{Wien}
\kurzfassung{Erstellung einer Echtzeit-Robotersteuerungsplattform unter Verwendung von Salamander OS, Xenomai, QEMU
und PCV-521 in der Yocto-Umgebung. Die Plattform basiert auf Salamander OS und nutzt Xenomai für Echtzeit-
Funktionen. Dazu muss im ersten Schritt die Virtualisierungsplattform evaluiert werden. (QEMU, Hyper-V, Virtual
Box, etc.) Als weiterer Schritt folgt die Anbindung eines Roboters über eine VARAN-Bus Schnittstelle. Das
gesamte System wird in der Yocto-Umgebung erstellt und konfiguriert.
Das Hauptziel der Arbeit ist es, herauszufinden, wie die Integration von Echtzeit-Funktionen und effizienten
Kommunikationssystemen in eine Robotersteuerungsplattform die Reaktionszeit und Zuverlässigkeit von
Roboteranwendungen verbessern kann}
\schlagworte{Schlagwort1, Schlagwort2, Schlagwort3, Schlagwort4}
\outline{Sections \ref{sec:salamander4-bare-metal} and \ref{sec:salamander4-virtualization}  demonstrate the inital real-time latency values gathered for bare metal and virtualization. 
}
\keywords{Echtzeit, Virtualisierung, Xenomai, VARAN}
%\acknowledgements{\blindtext}

@@

First, a comprehensive literature review is conducted to understand the current trends and challenges in real-time robot control. Based on the literature study, a suitable virtualization platform is selected.

After the selection of the virtualization platform, the robot control platform is implemented. This step includes the installation and configuration of Salamander OS, Xenomai, QEMU and PCV-521 in the Yocto environment. Once the platform has been implemented, the robot is connected via a VARAN bus interface. Finally, the platform is evaluated to determine how the integration of real-time functions and efficient communication systems improves the response time and reliability of robot applications.

@@

\section{Xenomai}
\noindent Xenomai 3 \cite{XenomaiXenomai} is a real-time framework that offers two paths to real-time performance. The first approach supplements the Linux kernel with a compact real-time core dubbed Cobalt, demonstrated in Figure \ref{fig:cobalt}. Cobalt runs side-by-side with Linux, but it handles all time-critical activities like interrupt processing and real-time thread scheduling with higher priority than the regular kernel activities.

@@

As a starting point, initial latency values of both the bare metal and virtualization versions were measured with the latency tool of the xenomai tool suite. Salamander 4 bare metal refers to the proprietary hardware of SIGMATEK used to employ the custom operating system. Salamander 4 virtualization refers to a virtual version of the Salamander 4 hardware platform, achieved through QEMU/KVM. Sections \ref{sec:salamander4-bare-metal} and \ref{sec:salamander4-virtualization} specify the details of the measurements for both versions. In the further course, the aim was to bring the latency values of the virtualization closer to those of the hardware and guarantee deterministic and reliable behavior.

@@

\noindent As a reference point, the latency program was executed on Salamander 4 bare metal for a duration of 10 minutes. The complete command used was \texttt{latency -h -g gnuplot.txt -T 600}, which runs the latency measurement tool for 600 seconds and prints histograms of min, avg, max latencies in a Gnuplot-compatible format to the file \texttt{gnuplot.txt}. Figure \ref{fig:max_latency_hardware} shows the gathered latency values in microseconds.

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/max_latency_hardware.png}
	\caption[Latency of Salamander 4 bare metal]{Latency of Salamander 4 bare metal}
	\label{fig:max_latency_hardware}
\end{figure}

Figure \ref{fig:gnuplot_max_latency_hardware} depicts the variation in latency over the course of said time. Since the data varies strongly, a logarithmic scale was used for both axes.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/0hardware/gnuplot_max_latency_hardware.png}
	\caption[Variation in latency of Salamander 4 bare metal]{Variation in latency of Salamander 4 bare metal}
	\label{fig:gnuplot_max_latency_hardware}
\end{figure}

@@

\begin{table}[h]
	\centering
	\caption{Latency Statistics in microseconds}
	\label{tab:latency_statistics}
	\begin{tabular}{|c|c|}\hline
	\textbf{Statistic} & \textbf{Value (\textmu s)} \\\hline
	Average Latency & 4.06 \\\hline
	Maximum Latency & 10.71 \\\hline
	Minimum Latency & 2.82 \\\hline
	Standard Deviation & 0.85 \\\hline
	\end{tabular}
	\end{table}

\section{Salamander 4 Virtualization}\label{sec:salamander4-virtualization}
In addition to providing Salamander 4 on its own hardware, SIGMATEK has also developed a virtualised version of this operating system. It was developed using Yocto, an open-source project that allows customised Linux distributions to be created for embedded systems \cite{WelcomeYoctoProject}. Upon generating the necessary files, Yocto provides a QEMU folder with the following components shown in code \ref{lst:ls}. QEMU is the environment in which the virtualization runs, as it is an open-source tool for hardware virtualization \cite{QEMU}.

@@

\noindent With the help of the script depicted in code \ref{script:qemu_def}, Salamander 4 is started together with the necessary hardware components in the QEMU environment. This makes it possible to run Salamander 4 on a variety of host systems, regardless of the specific hardware of the host. The following is a description of the components used for the virtualization of Salamander 4:
\begin{itemize}
	\item \textbf{bzImage}: Compressed Linux kernel image that is loaded by QEMU at system start. "bz" stands for big-zipped.
	\item \textbf{drive-c}: Directory serving as C drive for QEMU system, created and filled by qemu\_def.sh script.
	\item \textbf{ovmf.code.qcow2}: Firmware file for QEMU that enables UEFI boot process. \texttt{OVMF} stands for Open Virtual Machine Firmware, \texttt{qcow2} is a format for disk image files used by QEMU, it stands for "QEMU Copy On Write version 2".
	\item \textbf{qemu\_def.sh}: Shell script that starts QEMU with required parameters to boot Salamamder 4 OS. It is described in Code \ref{script:qemu_def}.
	\item \textbf{salamander-image-sigmatek-core2.ext4}: Disk image of the Salamander 4 OS for the Sigmatek Core 2 platform. It uses the \texttt{ext4} file system and serves as the root file system in the QEMU virtual machine, acting as the virtual hard drive.
	\item \textbf{stek-drive-c-image-sigmatek-core2.tar.gz}: Compressed tarball containing a pre-configured environment for the Salamander 4 OS. It is unpacked and sets up the \texttt{drive-c/} directory with system and log files in the \texttt{qemu\_def.sh} script. 
	\item \textbf{vmlinux}: Uncompressed Linux kernel image, typically used for debugging.
\end{itemize}

\begin{comment}
	https://www.baeldung.com/linux/kernel-images
\end{comment}

\noindent The inital QEMU script after the custom Yocto build and the starting point for this work is shown in Code \ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot Salamamder 4 OS. It will be adjusted in chapter \ref{cha:real-time_tuning} in order to accompany real-time performance tunings.

\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualization},label={script:qemu_def}]
	#!/bin/sh

@@

	exec qemu-system-x86_64 -M pc,accel=kvm -kernel ./bzImage \
	-m 2048 -drive file=salamander-image-sigmatek-core2.ext4,format=raw,media=disk \
	-append "console=ttyS0 console=tty1 root=/dev/sda rw panic=1 sigmatek_lrt.QEMU=1 ip=dhcp rootfstype=ext4" \
	-net nic,model=e1000,netdev=e1000 -netdev bridge,id=e1000,br=nm-bridge \
	-fsdev local,security_model=none,id=fsdev0,path=drive-c -device virtio-9p-pci,id=fs0,fsdev=fsdev0,mount_tag=/mnt/drive-C \
	-device vhost-vsock-pci,guest-cid=3,id=vsock0 \
	-drive if=pflash,format=qcow2,file=ovmf.code.qcow2 \
	-no-reboot -nographic
\end{lstlisting}
\end{minipage}

\noindent This script is run on a generic Ubuntu 22.04.4 system, as mentioned previously in chapter \ref{cha:methodology}. The kernel version and other details are presented in Code \ref{output:uname_a_ub}, using the \texttt{uname -a} command.

@@

Figure \ref{fig:gnuplot_max_latency_default} depicts the variation in latency over the course of said time with default settings.

@@

\begin{table}[h]
	\centering
	\caption{Latency Statistics default in microseconds}
	\label{tab:x}
	\begin{tabular}{|c|c|}\hline
	\textbf{Statistic} & \textbf{Value (\textmu s)} \\\hline
	Average Latency & 46.22 \\\hline
	Maximum Latency & 707.62 \\\hline
	Minimum Latency & 15.59 \\\hline
	Standard Deviation & 52.13 \\\hline
	\end{tabular}
	\end{table}

\noindent Comparing these values to those of bare metal, it is evident that there is a significant initial gap in the statistics. A maximum latency of 707.62~\textmu s is not tolerable for the system and needs to be tuned.

\clearpage
\chapter{Real-Time Performance Tuning}\label{cha:real-time_tuning}
\section{BIOS Configurations}
BIOS is firmware used to perform hardware initialization during the booting process. Some BIOS settings can have a significant impact on the real-time performance of the system [2]. The configurations in Table 4 are tuned for real-time performance.

@@

\clearpage
\section{Guest OS Configurations}
\vspace{1em}
\begin{minipage}{\linewidth}
	\begin{lstlisting}[name={QEMU script for starting Salamander 4 virtualization},label={script:qemu_configuration_optimizations}]
    #!/bin/sh

@@

\chapter{To Include}\label{cha:to_include}
\section{Latency Comparison}
In the initial phase, a comparative latency analysis was conducted between the hardware version and the virtualized version of Salamander 4. For this purpose, the latency tool of the Xenomai test suite was used. The latency was measured under two conditions, idle and CPU-stressed. The goal was to optimize the latency of the virtualization of Salamander 4 OS to closely match that of the bare metal version.

@@

Figure \ref{fig:kvm_exit_vapic} shows \texttt{kvm\_exit} frequency with APIC virtualization. Comparing this to the previous Figures \ref{fig:kvm_exit_taskset} and \ref{fig:kvm_exits_default}, it can be observed that an APIC\_Write no longer occurs.
